\documentclass[12pt,signature,logo]{BYUstatletter}

\renewcommand{\baselinestretch}{1.00}

\begin{document}

\begin{myletter}
{}
{}
{Dear Editors,}
{Sincerely,}

Please consider publishing in the \emph{R Journal} our manuscript, ``jsr223: A
Java Platform Integration for R with Programming Languages Groovy, JavaScript,
JRuby, Jython, and Kotlin.'' This manuscript is an introduction to the R
package jsr223: a simplified Java platform integration for R. Java is one of
the most successful development platforms in computing history. Its popularity
continues as more programming languages, tools, and technologies target the
Java virtual machine. The jsr223 project enables R programmers to more readily
take advantage of the development surrounding the Java platform. It does so by
providing a user-friendly interface to five programming languages that can use
and extend Java libraries: Groovy, JavaScript, Jython, JRuby, and Kotlin. The
jsr223 package also features extensive and configurable data exchange that
supports major R and Java data structures. In all, these and other jsr223
features enable Java integrations to be developed and executed rapidly in R.
Our paper includes illustrative code examples, a light treatment regarding
package implementation, and a thorough software review that relates jsr223 to
relevant R packages, such as rJava.

We confirm that this work is original and is not under consideration for
publication elsewhere.  Thank you for your consideration.

\end{myletter}

\end{document}   
 
