% !TeX root = master.tex
\chapter{Proposal}

To complete the project requirement for a Masters degree in Statistics from BYU, I propose to do the following:

\begin{enumerate}
\item Complete and submit the R software packages \textbf{jsr223} and \textbf{jdx} to The Comprehensive R Archive Network (CRAN) for distribution. The \textbf{jdx} package is not mentioned in this prospectus, but it will be discussed in the technical sections of the completed paper. The \textbf{jdx} package provides the core data exchange functionality for \textbf{jsr223}. Both \textbf{jsr223} and \textbf{jdx} will include R documentation and a vignette.

\item Complete and submit the associated paper to the \href{http://journal.r-project.org}{R Journal} \\ (\href{http://journal.r-project.org}{http://journal.r-project.org}). The paper will include: i.) an extended software review with other relevant software packages; ii.) a vignette to demonstrate the package's functionality; and iii.) code examples for each of the supported languages.
\end{enumerate}

\noindent A list of items corresponding to each task is included on the following pages.

\clearpage
\section{Complete and submit R software packages}

\subsection{Required}

\begin{enumerate}
\item Finish integration and testing with Kotlin programming language.

\item Address error due to a Kotlin dependency when package loads.

\item Finish vignette (content taken almost verbatim from R Journal paper) and insert vignette links into documentation.

\item Create public GitHub repositories for both packages. Enable issue tracker.

\item Build and check \textbf{jdx} and \textbf{jsr223} using latest development version of R software.

\item Submit \textbf{jsr223} packages to CRAN for review.
\end{enumerate}

\subsection{Optional}

\begin{enumerate}
\item Create build scripts for both packages.
\end{enumerate}

\subsection{Speculative}

\begin{enumerate}
\item Provide functionality to download and install script engines on demand.
\end{enumerate}

\clearpage
\section{Complete and submit paper to R Journal}

\subsection{Required}

\begin{enumerate}
\item Update content to reflect latest version of \textbf{jsr223}.

\item Complete extended software review including R packages relating to JavaScript, Groovy, and Python.

\item Determine if claim about \textbf{JRI} in software review is correct.

\item Address Kotlin and Scala integrations. Both have bugs and usage issues that must be listed. Those seeking a complete Scala integration will be pointed towards \textbf{rscala}. Also, existing generic code samples may require compatibility qualifications relating to Kotlin and Scala.

\item Update explanations regarding array ordering for n-dimensional arrays.

\item Finish all citations.

\item Post sample code in Github repositories. Put link in paper.

\item Write conclusion.
\end{enumerate}

\subsection{Optional}

\begin{enumerate}
\item Write instructions to include \textbf{jsr223} in other packages.

\item Include sample code to download script engines via SBT, Maven, and Gradle build and management utilities.

\item Complete a Groovy use case code sample.
\end{enumerate}

\subsection{Speculative}

\begin{enumerate}
\item Include use case for Kotlin language integration.

\item Complete known issues and limitations section. This would probably be better suited to Github or a living version of the vignette.
\end{enumerate}

